---
title: "Project Proposal"
date: "2025-11-06"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---
```{r, include=FALSE}
library(tidyverse)
library(haven)
```

### The group members (names and UNIs)
Disheng Jiang (dj2763), Yurou Liu (yl6109), Ruhui Shen (rs4788), Juan Tang (jt3649)

### The tentative project title
Exploring whether coffee is beneficial or harmful to human health by using the NHANES datasets from the CDC.

### The motivation for this project
Coffee, as one of the most popular beverages in the world, is highly favored by people. Since the caffeine in coffee has a regulatory effect on the central nervous system(Herqutanto et al., 2024), coffee is widely consumed for alertness. However, yet evidence on benefits and side effects (e.g., sleep disturbance, blood pressure) is mixed. 

The National Health and Nutrition Examination Survey (NHANES) is uniquely well-suited for this study. NHANES contains nationally representative sampling with detailed, standardized measurements that are directly relevant to caffeine. It also contains clinically measured outcomes like blood pressure, pulse and BMI, meanwhile providing rich covariates such as age, sex, physical activity, and socioeconomic markers. By utilizing the abundant raw data, we can learn about the effect of coffee on human health and dose-response patterns through data analysis and draw transparent and reproducible public health inferences.

### The intended final products

### The anticipated data sources
The project will use publicly available data from the 2021-2023 National Health and Nutrition Examination Survey (NHANES), maintained by the Centers for Disease Control and Prevention (CDC).

We will combine multiple datasets that recall participants’ demographic information, physical and laboratory examination results, and their dietary intake of coffee.

The demographic file (DEMO_L.xpt) provides key covariates such as age, sex, and socioeconomic indicators. Health outcome data will be drawn from laboratory files measuring metabolic, cardiovascular, and inflammatory biomarkers, including fasting glucose (GLU_L.xpt), insulin (INS_L.xpt), glycohemoglobin (GHB_L.xpt), cholesterol and triglycerides (TCHOL_L.xpt, HDL_L.xpt, TRIGLY_L.xpt), high-sensitivity C-reactive protein (HSCRP_L.xpt), and kidney and liver function markers (ALB_CR_L.xpt, BIOPRO_L.xpt). These datasets will be merged by participant ID (SEQN) and linked with dietary recall data capturing caffeine and coffee intake. 

Through statistical modeling and visualization, this combined dataset will allow assessment of potential health effects associated with coffee consumption.

### The planned analyses / visualizations / coding challenges

### The planned timeline
```{r, echo=FALSE}
timeline = tibble(
  date = c(
    "11/7",
    "11/8–11/12",
    "11/12–11/16",
    "11/15–11/16",
    "11/16–11/19",
    "11/19–11/20",
    "11/21–11/26",
    "11/26–12/1",
    "12/2–12/3",
    "12/3–12/6"
  ),
  work = c(
    "Determine project topic, find data source, complete the proposal",
    "Zoom meeting: determine final datasets and how to analyze data",
    "Complete final source collection",
    "Zoom meeting: determine detailed division of work",
    "Clean and tidy data",
    "Zoom meeting: design website structure",
    "Plot graphs and analyze data",
    "Complete report with visualizations and interpretations",
    "Proofread report",
    "Final webpage improvement and screencast"
  )
)

timeline |>
  knitr::kable(
    caption = "Project Timeline",
    align = c("c", "l")
  )
```

