---
title: "Project Proposal"
date: "2025-11-06"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---
```{r}
library(tidyverse)
library(haven)
```

### The group members (names and UNIs)
Disheng Jiang (dj2763), Yurou Liu (yl6109), Ruhui Shen (rs4788), Juan Tang (jt3649)

### The tentative project title
Exploring whether coffee is beneficial or harmful to human health by using the NHANES datasets from the CDC.

### The motivation for this project
Coffee, as one of the most popular beverages in the world, is highly favored by people. Since the caffeine in coffee has a regulatory effect on the central nervous system(Herqutanto et al., 2024), coffee is widely consumed for alertness. However, yet evidence on benefits and side effects (e.g., sleep disturbance, blood pressure) is mixed. 

The National Health and Nutrition Examination Survey (NHANES) is uniquely well-suited for this study. NHANES contains nationally representative sampling with detailed, standardized measurements that are directly relevant to caffeine. It also contains clinically measured outcomes like blood pressure, pulse and BMI, meanwhile providing rich covariates such as age, sex, physical activity, and socioeconomic markers. By utilizing the abundant raw data, we can learn about the effect of coffee on human health and dose-response patterns through data analysis and draw transparent and reproducible public health inferences.

### The intended final products

### The anticipated data sources

### The planned analyses / visualizations / coding challenges

### The planned timeline
```{r}
timeline = tibble(
  date = c(
    "11/7",
    "11/8–11/12",
    "11/12–11/16",
    "11/15–11/16",
    "11/16–11/19",
    "11/19–11/20",
    "11/21–11/26",
    "11/26–12/1",
    "12/2–12/3",
    "12/3–12/6"
  ),
  work = c(
    "Determine project topic, find data source, complete the proposal",
    "Zoom meeting: determine final datasets and how to analyze data",
    "Complete final source collection",
    "Zoom meeting: determine detailed division of work",
    "Clean and tidy data",
    "Zoom meeting: design website structure",
    "Plot graphs and analyze data",
    "Complete report with visualizations and interpretations",
    "Proofread report",
    "Final webpage improvement and screencast"
  )
)

timeline |>
  knitr::kable(
    caption = "Project Timeline",
    align = c("c", "l")
  )
```

